Efficiency

### 2018-02-04 DraftExec.Rnw
fix
with the battery discharge qualification

\usepackage[backend=bibtex, sorting=none, citestyle=authoryear]{biblatex}
\usepackage[margin=0.8in]{geometry}
\bibliography{references}

@@

\usepackage{draftwatermark}
\SetWatermarkText{DRAFT}
\SetWatermarkScale{2}
\SetWatermarkLightness{0.95}

@@

The selection of the DCNS bid to design the future submarine attracted considerable attention and controversy on a number of fronts. The choice of the pumpjet, and the question of efficiency at low speeds, has been touched on briefly in various informed discussions \parencite{stanford2017} \parencite{davies2017}.  However, it has remained the case that even extremely well-informed commentators reluctant to quantify and qualify the plausible scale of this trade-off, and what it might mean strategically and tactically for the submarine. Everyone knew it was an issue, but no-one confident enough in the scale of the impact (or even at transit speeds the direction of the impact \parencite{davies2017}) to make it clear whether it mattered. Even the head of the progam to build the submarine acknowledged an improvement in dived endurance would be possible with conventional propellers \parencite{sbs2017}, which seemed a clear acknowledgment that some trade-off was acknowledged, even by the strongest advocates of the pumpjet.  But despite this, the chiefs of DCNS and SEA 1000 both remain adamant that the pumpjet was an "integral part of our bid" \parencite{guillou2017} and that it can be efficient "across the entire speed range" \parencite{sammut2017}.

Having done my higher education in physics, and spent considerable time considering military maritime technology running a small start-up defence company I found seemingly unresolved series of exchanges regarding the pumpjet choice particulaly curious.  As is commonly quipped, including in defence circles, it can be reliably assumed when it comes to technology choices "there's no free lunch in physics".  Whilst all technologies steadily improve, and new ones continuously emerge, it's inevitably the case that when significant technology switches are made, some trade-offs are necessary. In the case of the pumpjets, the trade-offs what might have been expected appeared obvious, and probably unavoidable to some extent to all those experienced in maritime systems: it would have lower propulsive efficiency, particularly at low speeds.

It's definitely true that efficiency isn't everything in military technology.  In recent years I was heavily involved in designing a small speed-boat and a medium-weight torpedo for naval combat, and in both cases I selected a waterjet or pumpjet for those purpose. Pumpjets, or some kind of "decelerating ducted propeller" (explained in detail later) suited very well the sorts of speeds and operational circumstances that they were required to operate in, which were well over 30kt. However, having made that selection, and on several occasions wrestled with adapting the boat to alternative purposes, (such as mine warfare, including the towing of heavy sweeps and decoys) I found myself continually balancing the inclination to design a better, more specialised tool for each job, with the desire to compromise performance consolidate capability into a multi-role, more flexible capability.  The trade-offs inherent in propulsion systems, and the relative efficency of jets at high speeds and propellers at low speeds was a familiar encounter.

@@

The choice of the pumpjet propulsion system is quite probably the most significant new technology inclusion for Australia's future submarines.  Unlike other modern technolgoies such as Air Independent Propulsion and Lithium Ion Batteries, which are either in service or entering service in multiple other navies, no other navies have have adopted pumpjet technology for their conventional sumbarines, which is currently only used by nuclear submarines \parencite(coorey2016).  Questions have been raised about the efficiency of pumpjets at low speeds \parencite{davies2017}. Propulsive efficiency at low speeds might be very important for conventional submarines since it will largely determine how long they can remain underwater before sufracing to recharge their batteries, but it would be of little or no concern to a nuclear submarine, which has effectively unlimited energy stores.

Despite acknowledging that the pumpjet may have an adverse impact on dived endurance \parencite{sbs2017}, Naval Group justifies the superiority of the pumpjet in the following terms: "In a contest between two otherwise identical submarines, the one with pump jet propulsion always has the tactical advantage."\parencite{guillou2017}  No further explanation of this superiority has been provided in detail.  This report does not attempt to engage with specific tactical encounters such as this, however, it is widely held that the pumpjet offered acoustic advantages which were highly regarded by those Australians who were responsible for selecting the design partner \parencite{stewart2016}.  Commentary from the designers of Shortfin Barracuda Block 1A seem to confirm that it is an acoustic advantage at elevated speeds which drive this tactical advantage when facing another submarine in combat \parencite{autret2016}.

The comparison between two "otherwise identical" submarines (though not necessarily limited to engaging just each other in combat), provides a useful and suitable framework for evaluating a potential trade-off in terms of other parameters, such as propulsive efficiency.  This is particularly the case given that most other significantly novel or new technologies which might dramatically impact propulsive performance were apparently not included for serious consideration \parencite[10]{stanford2017}.  Whilst many of the pages of this report are concerned with establishing the probable magnitude of the impact of efficiency at a range of speeds, the existence and direction of an efficiency difference between the two systems was quite firmly established at all the speeds most important for sumbarine operations. This allows for a surprisingly simple and strong qualitative conclusion to be reached in terms of the "otherwise identical" submarine comparison.

@@

This report also explores in some detail the theoretical concepts that drive the limitations of pumpjet performance at lower speeds in comparison to propellers.  Enduring and fundamental principals of momentum theory provide robust limitations on the circumstances where a propeller's 'ideal efficiency' might be improved by the addition of a duct/or shroud, one type of which constitutes the transition to a 'pumpjet'.  These principals stronly support the the notion that pumpjets will not and indeed cannot suddenly overcome their efficiency disadvantage at low speeds by means of technical ingenuity or innovation.

Furthmore, a particular hydrodynamic phenonemon called 'flow separation' can be identified as the primary cause of the pumpjet's declining efficiency at very low speeds. Put simply, the duct provides an additional hurdle or barrier though which the water must pass when moving through the propulsor.  At very low waterspeeds, this additional resistance is sufficient to stop some of the flow progressing through the jet, creating unsteady eddies and vortices which sap energy from the system, but produce no thrust, driving down efficiency to zero at very low flow-rates.

@@

\caption{(a) At high speeds, the intake of a pumpjet slows the waterflow and raises its pressure, and it subsequently accelerates through the contracting nozzle to form an efficient jet (b) At low speeds, the pressure gradient created by the duct causes some of the waterflow to stop and reverse.  This 'flow separation' wastes energy spinning water in circles, and dramatically reduces the system's efficiency.}
\label{fig:HiLoSmall.png}
\end{figure}

@@

\begin{tabular}{ |p{3cm}|p{11cm}| }
 \hline
 \multicolumn{2}{|c|}{Assumption Table for "otherwise identical" submarines} \\
 \hline
 Batteries & 500 tonnes lead-acid, 0.14MJ/kg density, 100\% discharge\\
 Power Reference & 20kt reached by pumpjet submarine with 6.5MW power delivered\\
 Efficiency Curve & Central lines in Figure \ref{fig:Efficiency_assumptions} modeled on those in Figure \ref{fig:BMTEfficiency.png}\\
 Diesel Generators & 5MW delivered for effective re-charging\\
 Hotel Load & Constant 100kW\\
 AIP Technology & None\\
 \hline
\end{tabular}

@@

\begin{tcolorbox}[width=0.8\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= Quick Summary: Sped and Drag]
Drag directly determines the energy required to cover a given distance. The physics relevant to submarines require that drag increases with speed squared, and the power required by a vessel to reach a given speed scales with speed cubed.  Consequently, the energy \textit{required} to move a vessel a given distance can vary massively over different speeds, often much more than the efficiency of any given system \textit{delivering that energy} will tend to vary.
\\
\\
This means that most vessels will be able to go much further by travelling extremely slowly, almost regardless of particulars of just how they are designed and what propulsion systems they use.  It also means that almost all the attention of propulsion system design focuses on 'design speed', generally pretty close to the top speed, since at very low speed so little thrust (hence fuel) is required that efficiency is of little concern.
\\
\\
For conventional submarines, however, the situation is very different.  They already adopt very low speeds (8-10kt) for transit to maximise the range they can get from their diesel.  Even most commercial shipping, including bulk carriers generally go much faster (16-24kt).  When a submarine dives, it relies on its batteries, which might store less than 1\% of the energy that's stored in its diesel tanks, and consequently has to slow down much more again (~4kt) to get meaningful range before returning to the surface to recharge. Whilst much higher speeds are possible even when dived, up to ~20kt, this has such a dramatic impact on the power demanded that it can't be sustained for useful periods of time, and are consequently only useful in an emergency, or very particular rapidly evolving tactical circumstances.
\\
\\
Since a conventional submarine already has to slow down to a crawl to get good dived range and endurance, the efficiency of the propulsion system now becomes of very great importance, as poor efficiency here will have a signifcant impact on how soon they have to return to the surface.  Nuclear submarines have effectively unlimited energy stores and don't require air to use it, and consequently don't share this concern.  They might also also spend much more of their dived time at much higher speeds as a result.
\end{tcolorbox}

\subsection{Practical implications for submarines}
This has a profound impact on the operation and engineering of maritime vessels.  Reducing velocity has such a substantial reduction on hull drag that going slower is almost always a reasonable means of conserving fuel overall.  To the extent that time is non-critical, slower is always much much better.  It's for this reason that during the financial crisis, many cargo shipping companies adopted the practice of slow steaming \parencite{liang2014} in order to conserve fuel, despite this causing a range of possible new engineering issues which need to be accounted for in order to operate the engines at lower than normal power for sustained periods \parencite[8-10]{sanguri2012}. The fuel saving from operating even 30\% slower means that around half the total energy is required for a given journey.  Other incremental costs and inefficiences from operating 'off-design' are frequently outweight by such a dramatic reduction in overall power demand.

@@

However, the difference between the peak-powers of the submarines significantly understates how different the designs are, because the difference in the total amount of energy stored and available for use in a voyage or dive is vastly greater.  A nuclear submarine might have literally millions of times more energy at its disposal, it is widely regarded as practically unlimited for all intents and purposes on a given voyage. Consequently, wheras a nuclear submarine might regularly conduct transits at or near it's peak power, a diesel electric submarine would probably transit at less than half the speed which it could manage at a sprint, (and use less than a quarter of the energy, as discussed earlier) and on patrol it might be operating at a tenth of the maximum speed, and use maybe just 1\% of as much power on propulsion. In this situation, the amount of power that is drawn for lights, CO2 scrubbing, washing, cooking, heating, as well as the electronics driving the combat systems (the 'Hotel Load') might well become significant, and even be as larger or larger than what is required for propulsion.

@@

\subsection{Xenon poisoning and low power limitations}
Perhaps more significantly for power-design of nuclear submarines is a little-discussed phenomenon called Xenon poisoning, which affects nuclear reactors when they shut down or lower their power significantly. When Uranium or Plutonium atoms split (or fission) into two smaller isotopes atoms, a variety of different radio nucleides (unstable variants of atoms) are produced.  Two of these are Xenon-135 and Iodine 135.  Iodine is produced much more often, and decays quickly into Xenon-135 (which has a half-life of about nine hours), and it is this process which produces the majority of Xenon in the reactor.  This Xenon has a very special, perhaps unique role in reactors, since they very easily captures the free neutrons which trigger the continued fission reactions of Uranium or Plutonium which drive the reactor.  As such, Xenon is known as reactor 'poison' since it can kill the reactor's reactivity in very high doses.

This high neutron-capture from Xenon means that it has a duplicitous relationship with the reactor's power level.  At high levels of reactor power, lots of Xenon-135 and Iodine-135 are produced by the fission process.  The high presence of Xenon reduced the reactor's reactivity.  But another process works in the opposite direction which offsets this, since there are lots of neutrons available to 'burn off' the Xenon (which absobs the neturons to become a different isotope).  This keeps the total level of Xenon in check.

The situation changes suddenly when the reactor undergoes a sudden change in power output.  If the power is lowered dramatically and suddenly, the production of Xenon continues quite rapidly for some time due to the decay of the large stock of Iodine-135.  With less neutron flux available to 'burn off' the Xenon, the Xenon levels spike, and push down the reactor's reactivity.  Unless the reactor is quickly raised back to relatively high power (~60\%) quite quickly (an hour or less), the Xenon levels become so high that the reactor will have to be shut down, otherwise extreme (and dangerous) measures would be required in order to keep the reactor going.  (This is essentially what lead to the Chernobyl Explosion \parencite{WNA2009}.)  A fuller discussion of Xenon poisoning can be found in \cite{garland2005}, which demonstrates key concepts related to the poisoning effect shown in Figure \ref{fig:XenonPoison.png}.

@@

It should also be noted that a nuclear reactor's aversion to sudden reductions in power would also have a substantial impact on the design of a submarine's combat system, and it's demand on the Hotel Load. For the same reason, a high Hotel Load, or power-hungry Combat System, could actually be advantageous, as it helps to set an elevated 'floor' for power requirements, reducing the scale of fluctuations in overall power demand from the reactor due to changes in propulsion speed.

\begin{tcolorbox}[width=0.8\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= The other reason reactors are different]
Nuclear reactors have an enormous amount of energy stored in a small space, so for submarines they may not need refuelling for many years, if every.  However, they vary from other power plants in another crucial way, that can have downsides to their operations.
\\
\\
Most people take an internal combustion engine's ability to vary its power output for granted.  Simply open the throttle, and more power is produced, ease it right off and it drops again.  The engines that drive our cars and trucks are used to going from full power to almost no power every time we change gears, or stop in traffic.  Other than a tiny bit more long-term wear on the engine, there's little consequence to large power fluctuations.
\\
\\
Nuclear reactors don't enjoy the same luxury.  The fissioning of Uranium or Plutonium atoms leaves a range of other unstable nuclear isotopes as by-products, which in turn decay to other nuclear isotopes, some of which absorb neutrons in the reactor core, or have other impacts on the 'reactivity' of the reactor (how fast it goes). As a consequence, nuclear reactors can't freely oscilate thier power levels, as some of these by-products following sudden changes (particulalry dramatic reductions in power) can produce dangerous situations where the reactor can become unstable.  (It was following such a exercise in reducing power that the Chernobyl explosion occurred.)
\\
\\
The physics of these decay chains mean that nuclear reactors, including those in submarines generally avoid sudden changes in power.  Large reductions in power have to be taken in staged steps or a steady ramp-down over many hours, or more likely a couple of days.  Furthermore, if the reactor is ever completely shut-down, it generally can't be started safely again for some time, most likely a couple of days.  For obvious reasons, an operating sumbarine will strictly avoid anything that might risk requiring a shutdown.
\\
\\
The result is that nuclear submarines whenever operating at reduced speeds tend to have an abundance of spare power to dispose of.  The normal desire for energy 'efficiencies' are thrown into reverse, since it's undesirable to reduce the power of the reactor very far, and dangerous to reduce it fast. Whatever excess energy isn't sucked up by hotel load and propulsion has to be dispersed by additional systems, generally pumping water to dispose of the excess energy as heat. As such, in nuclear sumbarine design trading-off propulsive efficiency at low speeds is an easy, or even desirable choice.
\end{tcolorbox}

\subsection{Recent Commentary}
Some efforts have been made to 'debunk' such suggestions that nuclear submarines effectively have "unlimited" power \parencite{autret2016}, with the leaders of the DCNS's bid at the time making some effort to suggest that the needs and characteristics of nuclear and diesel-electric submarines "drives the same system design". They make three comments to justify this claim.  The first is that the large submarine proposed by DCNS for Australia uses as much or more power for the hotel load at the most frequently used speeds.  Assuming the authors are referring to speeds between patrol speeds and approaching transit speed, the subsequent analysis will show that this fact may be true, particulalry in the presence of the pumpjet. However on its own this doesn't constitute an argument worthy of any consideration. One might equally argue a glutton can be relied upon not to eat too much at dinner because he eats a dinner sized meal at afternoon tea.

@@

Claims that the power management imperatives of nuclear submarines and diesel electric submarines are even remotely similar at low speeds are generally implausible.  Nuclear submarines will have unconstrained energy reserves, and a vast surplus of spare power capacity when ever operating below top speed.  The dynamics of reactor management dictate that generally avoiding large fluctuation in power, and strictly avoiding sharp and dramtic reductions, will be dominant considerations in a nuclear submarine's power output.  To put things simply, the characterisation that nuclear submarines have "power to burn" is generally sound, and would only ever be challenged when the submarine is travelling at flank speed, or the reactor is shut down due to a (presumeably near-catastrophic) emergency.

@@

The selection of the exact blade selection that is used to define pitch must be specified as being at some fraction of the radius from the centre, given the tendancy for the blades to twist.  This distance is calculated to be the 'moment mean' or a technical derived effective average, which tends to lie between 0.6R and 0.7R.  A thorough discussion of pitch can be found in \cite[35-37]{carlton2007}.

@@

\subsubsection{Pitch and the Advance Ratio}
Pitch is also an extremely important concept to understand because it relates closely to the way that a propeller's 'open-water efficiency' $\eta_O$ is expressed in charts, as well as the advance ratio $J$ that usually comprises the x-axis in such charts.

@@

It can be seen from the diagram of the open-water efficiencies that there is some desirable non-zero slip ratio for a propeller.  The other end of the spectrum, where efficiencies also fall off toward zero, represent circumstances where the propeller is simply spinning in the same place in the water. In this case the slip ratio is one.  This represents a circumstance known as 'bollard pull' where a vessel is simply exerting force against an immovable object or infinite load.  Whilst the very low efficiencies here could represent a case of a propeller that is simply too small for the task it is trying to achieve, it is actually a significant and important circumstance to consider for propellers on working vessels such as tugs or trawlers, which exert large forces at very low water-speeds.  Measurement of the thrust generated in bollard pull, and the power required to achieve this thrust, is still an important metric for such vessels, but the definition of 'useful work' for normal propulsion requieres the thust to be excerted over some distance travelled.

@@

A further means by which the efficiency of a propeller can be increased is through the reduction of the turbulence created by the blade-tips passing through the open (static) water at high speed.  Since the two sides of the blades have low and high pressure on them respectively, at the tips where the two faces meet, there there is a tendancy for water to 'spill' from the high pressure side the the low pressure side, creating a vortex which leads to unwanted turbulence.  It is also around this vortex that blade tip cavitation tends to occur. By limiting the movement of water around the outside of the blade tip by having a tightly fitted shroud, losses due to this particulary type of tubulence, and the onset of this type of cavitation, or the vortices that occur even when cavitation does not, can be diminished.

>>>>

\usepackage[backend=bibtex, sorting=none, citestyle=numeric]{biblatex}
\usepackage[margin=0.8in]{geometry}
\bibliography{references}

@@

\usepackage{draftwatermark}
\SetWatermarkText{DRAFT}
\SetWatermarkScale{1.5}
\SetWatermarkLightness{0.95}

@@

The selection of the DCNS bid to design the future submarine attracted considerable attention and controversy on a number of fronts. The choice of the pumpjet, and the question of efficiency at low speeds, has been touched on briefly in various informed discussions \parencite{stanford2017} \parencite{davies2017}.  However, it has remained the case that even extremely well-informed commentators reluctant to quantify and qualify the plausible scale of this trade-off, and what it might mean strategically and tactically for the submarine. Everyone knew it was an issue, but no-one confident enough in the scale of the impact (or even at transit speeds the direction of the impact \parencite{davies2017}) to make it clear whether it mattered. Even the head of the progam to build the submarine acknowledged an improvement in dived endurance would be possible with conventional propellers \parencite{sbs2017}, which seemed a clear acknowledgment that some trade-off was acknowledged, even by the strongest advocates of the pumpjet.  But despite this, the chiefs of DCNS and SEA 1000 both remain adamant that the pumpjet was an ``integral part of our bid" \parencite{guillou2017} and that it can be efficient ``across the entire speed range" \parencite{sammut2017}.

Having done my higher education in physics, and spent considerable time considering military maritime technology running a small start-up defence company I found seemingly unresolved series of exchanges regarding the pumpjet choice particulaly curious.  As is commonly quipped, including in defence circles, it can be reliably assumed when it comes to technology choices ``there's no free lunch in physics".  Whilst all technologies steadily improve, and new ones continuously emerge, it's inevitably the case that when significant technology switches are made, some trade-offs are necessary. In the case of the pumpjets, the trade-offs what might have been expected appeared obvious, and probably unavoidable to some extent to all those experienced in maritime systems: it would have lower propulsive efficiency, particularly at low speeds.

It's definitely true that efficiency isn't everything in military technology.  In recent years I was heavily involved in designing a small speed-boat and a medium-weight torpedo for naval combat, and in both cases I selected a waterjet or pumpjet for those purpose. Pumpjets, or some kind of ``decelerating ducted propeller" (explained in detail later) suited very well the sorts of speeds and operational circumstances that they were required to operate in, which were well over 30kt. However, having made that selection, and on several occasions wrestled with adapting the boat to alternative purposes, (such as mine warfare, including the towing of heavy sweeps and decoys) I found myself continually balancing the inclination to design a better, more specialised tool for each job, with the desire to compromise performance consolidate capability into a multi-role, more flexible capability.  The trade-offs inherent in propulsion systems, and the relative efficency of jets at high speeds and propellers at low speeds was a familiar encounter.

@@

The choice of the pumpjet propulsion system is quite probably the most significant new technology inclusion for Australia's future submarines.  Unlike other modern technolgoies such as Air Independent Propulsion and Lithium Ion Batteries, which are either in service or entering service in multiple other navies, no other navies have have adopted pumpjet technology for their conventional sumbarines, which is currently only used by nuclear submarines \parencite{coorey2016}.  Questions have been raised about the efficiency of pumpjets at low speeds \parencite{davies2017}. Propulsive efficiency at low speeds might be very important for conventional submarines since it will largely determine how long they can remain underwater before sufracing to recharge their batteries, but it would be of little or no concern to a nuclear submarine, which has effectively unlimited energy stores.

Despite acknowledging that the pumpjet may have an adverse impact on dived endurance \parencite{sbs2017}, Naval Group justifies the superiority of the pumpjet in the following terms: ``In a contest between two otherwise identical submarines, the one with pump jet propulsion always has the tactical advantage."\parencite{guillou2017}  No further explanation of this superiority has been provided in detail.  This report does not attempt to engage with specific tactical encounters such as this, however, it is widely held that the pumpjet offered acoustic advantages which were highly regarded by those Australians who were responsible for selecting the design partner \parencite{stewart2016}.  Commentary from the designers of Shortfin Barracuda Block 1A seem to confirm that it is an acoustic advantage at elevated speeds which drive this tactical advantage when facing another submarine in combat \parencite{autret2016}.

The comparison between two ``otherwise identical" submarines (though not necessarily limited to engaging just each other in combat), provides a useful and suitable framework for evaluating a potential trade-off in terms of other parameters, such as propulsive efficiency.  This is particularly the case given that most other significantly novel or new technologies which might dramatically impact propulsive performance were apparently not included for serious consideration \parencite[10]{stanford2017}.  Whilst many of the pages of this report are concerned with establishing the probable magnitude of the impact of efficiency at a range of speeds, the existence and direction of an efficiency difference between the two systems was quite firmly established at all the speeds most important for sumbarine operations. This allows for a surprisingly simple and strong qualitative conclusion to be reached in terms of the ``otherwise identical" submarine comparison.

@@

This report also explores in some detail the theoretical concepts that drive the limitations of pumpjet performance at lower speeds in comparison to propellers.  Enduring and fundamental principals of momentum theory provide robust limitations on the circumstances where a propeller's `ideal efficiency' might be improved by the addition of a duct/or shroud, one type of which constitutes the transition to a `pumpjet'.  These principals stronly support the the notion that pumpjets will not and indeed cannot suddenly overcome their efficiency disadvantage at low speeds by means of technical ingenuity or innovation.

Furthmore, a particular hydrodynamic phenonemon called `flow separation' can be identified as the primary cause of the pumpjet's declining efficiency at very low speeds. Put simply, the duct provides an additional hurdle or barrier though which the water must pass when moving through the propulsor.  At very low waterspeeds, this additional resistance is sufficient to stop some of the flow progressing through the jet, creating unsteady eddies and vortices which sap energy from the system, but produce no thrust, driving down efficiency to zero at very low flow-rates.

@@

\caption{(a) At high speeds, the intake of a pumpjet slows the waterflow and raises its pressure, and it subsequently accelerates through the contracting nozzle to form an efficient jet (b) At low speeds, the pressure gradient created by the duct causes some of the waterflow to stop and reverse.  This `flow separation' wastes energy spinning water in circles, and dramatically reduces the system's efficiency.}
\label{fig:HiLoSmall.png}
\end{figure}

@@

\begin{tabular}{ |p{3cm}|p{11cm}| }
 \hline
 \multicolumn{2}{|c|}{Assumption Table for ``otherwise identical" submarines} \\
 \hline
 Batteries & 500 tonnes lead-acid, 0.14MJ/kg density, 100\% discharge\\
 Power Reference & 20kt reached by pumpjet submarine with 6.5MW power delivered\\
 Efficiency Curve & Central lines in Figure \ref{fig:Efficiency_assumptions} modeled on those in Figure \ref{fig:BMTEfficiency.png}\\
 Diesel Generators & 5MW delivered for effective re-charging\\
 Hotel Load & Constant 100kW\\
 AIP Technology & None\\
 \hline
\end{tabular}

@@

\begin{tcolorbox}[width=0.8\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= Quick Summary: Sped and Drag]
Drag directly determines the energy required to cover a given distance. The physics relevant to submarines require that drag increases with speed squared, and the power required by a vessel to reach a given speed scales with speed cubed.  Consequently, the energy \textit{required} to move a vessel a given distance can vary massively over different speeds, often much more than the efficiency of any given system \textit{delivering that energy} will tend to vary.
\\
\\
This means that most vessels will be able to go much further by travelling extremely slowly, almost regardless of particulars of just how they are designed and what propulsion systems they use.  It also means that almost all the attention of propulsion system design focuses on `design speed', generally pretty close to the top speed, since at very low speed so little thrust (hence fuel) is required that efficiency is of little concern.
\\
\\
For conventional submarines, however, the situation is very different.  They already adopt very low speeds (8-10kt) for transit to maximise the range they can get from their diesel.  Even most commercial shipping, including bulk carriers generally go much faster (16-24kt).  When a submarine dives, it relies on its batteries, which might store less than 1\% of the energy that's stored in its diesel tanks, and consequently has to slow down much more again (~4kt) to get meaningful range before returning to the surface to recharge. Whilst much higher speeds are possible even when dived, up to ~20kt, this has such a dramatic impact on the power demanded that it can't be sustained for useful periods of time, and are consequently only useful in an emergency, or very particular rapidly evolving tactical circumstances.
\\
\\
Since a conventional submarine already has to slow down to a crawl to get good dived range and endurance, the efficiency of the propulsion system now becomes of very great importance, as poor efficiency here will have a signifcant impact on how soon they have to return to the surface.  Nuclear submarines have effectively unlimited energy stores and don't require air to use it, and consequently don't share this concern.  They might also also spend much more of their dived time at much higher speeds as a result.
\end{tcolorbox}

\subsection{Practical implications for submarines}
This has a profound impact on the operation and engineering of maritime vessels.  Reducing velocity has such a substantial reduction on hull drag that going slower is almost always a reasonable means of conserving fuel overall.  To the extent that time is non-critical, slower is always much much better.  It's for this reason that during the financial crisis, many cargo shipping companies adopted the practice of slow steaming \parencite{liang2014} in order to conserve fuel, despite this causing a range of possible new engineering issues which need to be accounted for in order to operate the engines at lower than normal power for sustained periods \parencite[8-10]{sanguri2012}. The fuel saving from operating even 30\% slower means that around half the total energy is required for a given journey.  Other incremental costs and inefficiences from operating `off-design' are frequently outweight by such a dramatic reduction in overall power demand.

@@

However, the difference between the peak-powers of the submarines significantly understates how different the designs are, because the difference in the total amount of energy stored and available for use in a voyage or dive is vastly greater.  A nuclear submarine might have literally millions of times more energy at its disposal, it is widely regarded as practically unlimited for all intents and purposes on a given voyage. Consequently, wheras a nuclear submarine might regularly conduct transits at or near it's peak power, a diesel electric submarine would probably transit at less than half the speed which it could manage at a sprint, (and use less than a quarter of the energy, as discussed earlier) and on patrol it might be operating at a tenth of the maximum speed, and use maybe just 1\% of as much power on propulsion. In this situation, the amount of power that is drawn for lights, CO2 scrubbing, washing, cooking, heating, as well as the electronics driving the combat systems (the `Hotel Load') might well become significant, and even be as larger or larger than what is required for propulsion.

@@

\subsection{Xenon poisoning and low power limitations}
Perhaps more significantly for power-design of nuclear submarines is a little-discussed phenomenon called Xenon poisoning, which affects nuclear reactors when they shut down or lower their power significantly. When Uranium or Plutonium atoms split (or fission) into two smaller isotopes atoms, a variety of different radio nucleides (unstable variants of atoms) are produced.  Two of these are Xenon-135 and Iodine 135.  Iodine is produced much more often, and decays quickly into Xenon-135 (which has a half-life of about nine hours), and it is this process which produces the majority of Xenon in the reactor.  This Xenon has a very special, perhaps unique role in reactors, since they very easily captures the free neutrons which trigger the continued fission reactions of Uranium or Plutonium which drive the reactor.  As such, Xenon is known as reactor `poison' since it can kill the reactor's reactivity in very high doses.

This high neutron-capture from Xenon means that it has a duplicitous relationship with the reactor's power level.  At high levels of reactor power, lots of Xenon-135 and Iodine-135 are produced by the fission process.  The high presence of Xenon reduced the reactor's reactivity.  But another process works in the opposite direction which offsets this, since there are lots of neutrons available to `burn off' the Xenon (which absobs the neturons to become a different isotope).  This keeps the total level of Xenon in check.

The situation changes suddenly when the reactor undergoes a sudden change in power output.  If the power is lowered dramatically and suddenly, the production of Xenon continues quite rapidly for some time due to the decay of the large stock of Iodine-135.  With less neutron flux available to `burn off' the Xenon, the Xenon levels spike, and push down the reactor's reactivity.  Unless the reactor is quickly raised back to relatively high power (~60\%) quite quickly (an hour or less), the Xenon levels become so high that the reactor will have to be shut down, otherwise extreme (and dangerous) measures would be required in order to keep the reactor going.  (This is essentially what lead to the Chernobyl Explosion \parencite{WNA2009}.)  A fuller discussion of Xenon poisoning can be found in \cite{garland2005}, which demonstrates key concepts related to the poisoning effect shown in Figure \ref{fig:XenonPoison.png}.

@@

It should also be noted that a nuclear reactor's aversion to sudden reductions in power would also have a substantial impact on the design of a submarine's combat system, and it's demand on the Hotel Load. For the same reason, a high Hotel Load, or power-hungry Combat System, could actually be advantageous, as it helps to set an elevated `floor' for power requirements, reducing the scale of fluctuations in overall power demand from the reactor due to changes in propulsion speed.

\begin{tcolorbox}[width=0.8\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= The other reason reactors are different]
Nuclear reactors have an enormous amount of energy stored in a small space, so for submarines they may not need refuelling for many years, if every.  However, they vary from other power plants in another crucial way, that can have downsides to their operations.
\\
\\
Most people take an internal combustion engine's ability to vary its power output for granted.  Simply open the throttle, and more power is produced, ease it right off and it drops again.  The engines that drive our cars and trucks are used to going from full power to almost no power every time we change gears, or stop in traffic.  Other than a tiny bit more long-term wear on the engine, there's little consequence to large power fluctuations.
\\
\\
Nuclear reactors don't enjoy the same luxury.  The fissioning of Uranium or Plutonium atoms leaves a range of other unstable nuclear isotopes as by-products, which in turn decay to other nuclear isotopes, some of which absorb neutrons in the reactor core, or have other impacts on the `reactivity' of the reactor (how fast it goes). As a consequence, nuclear reactors can't freely oscilate thier power levels, as some of these by-products following sudden changes (particulalry dramatic reductions in power) can produce dangerous situations where the reactor can become unstable.  (It was following such a exercise in reducing power that the Chernobyl explosion occurred.)
\\
\\
The physics of these decay chains mean that nuclear reactors, including those in submarines generally avoid sudden changes in power.  Large reductions in power have to be taken in staged steps or a steady ramp-down over many hours, or more likely a couple of days.  Furthermore, if the reactor is ever completely shut-down, it generally can't be started safely again for some time, most likely a couple of days.  For obvious reasons, an operating sumbarine will strictly avoid anything that might risk requiring a shutdown.
\\
\\
The result is that nuclear submarines whenever operating at reduced speeds tend to have an abundance of spare power to dispose of.  The normal desire for energy `efficiencies' are thrown into reverse, since it's undesirable to reduce the power of the reactor very far, and dangerous to reduce it fast. Whatever excess energy isn't sucked up by hotel load and propulsion has to be dispersed by additional systems, generally pumping water to dispose of the excess energy as heat. As such, in nuclear sumbarine design trading-off propulsive efficiency at low speeds is an easy, or even desirable choice.
\end{tcolorbox}

\subsection{Recent Commentary}
Some efforts have been made to `debunk' such suggestions that nuclear submarines effectively have ``unlimited" power \parencite{autret2016}, with the leaders of the DCNS's bid at the time making some effort to suggest that the needs and characteristics of nuclear and diesel-electric submarines ``drives the same system design". They make three comments to justify this claim.  The first is that the large submarine proposed by DCNS for Australia uses as much or more power for the hotel load at the most frequently used speeds.  Assuming the authors are referring to speeds between patrol speeds and approaching transit speed, the subsequent analysis will show that this fact may be true, particulalry in the presence of the pumpjet. However on its own this doesn't constitute an argument worthy of any consideration. One might equally argue a glutton can be relied upon not to eat too much at dinner because he eats a dinner sized meal at afternoon tea.

@@

Claims that the power management imperatives of nuclear submarines and diesel electric submarines are even remotely similar at low speeds are generally implausible.  Nuclear submarines will have unconstrained energy reserves, and a vast surplus of spare power capacity when ever operating below top speed.  The dynamics of reactor management dictate that generally avoiding large fluctuation in power, and strictly avoiding sharp and dramtic reductions, will be dominant considerations in a nuclear submarine's power output.  To put things simply, the characterisation that nuclear submarines have ``power to burn" is generally sound, and would only ever be challenged when the submarine is travelling at flank speed, or the reactor is shut down due to a (presumeably near-catastrophic) emergency.

@@

The selection of the exact blade selection that is used to define pitch must be specified as being at some fraction of the radius from the centre, given the tendancy for the blades to twist.  This distance is calculated to be the `moment mean' or a technical derived effective average, which tends to lie between 0.6R and 0.7R.  A thorough discussion of pitch can be found in \cite[35-37]{carlton2007}.

@@

\subsubsection{Pitch and the Advance Ratio}
Pitch is also an extremely important concept to understand because it relates closely to the way that a propeller's `open-water efficiency' $\eta_O$ is expressed in charts, as well as the advance ratio $J$ that usually comprises the x-axis in such charts.

@@

It can be seen from the diagram of the open-water efficiencies that there is some desirable non-zero slip ratio for a propeller.  The other end of the spectrum, where efficiencies also fall off toward zero, represent circumstances where the propeller is simply spinning in the same place in the water. In this case the slip ratio is one.  This represents a circumstance known as `bollard pull' where a vessel is simply exerting force against an immovable object or infinite load.  Whilst the very low efficiencies here could represent a case of a propeller that is simply too small for the task it is trying to achieve, it is actually a significant and important circumstance to consider for propellers on working vessels such as tugs or trawlers, which exert large forces at very low water-speeds.  Measurement of the thrust generated in bollard pull, and the power required to achieve this thrust, is still an important metric for such vessels, but the definition of `useful work' for normal propulsion requieres the thust to be excerted over some distance travelled.

@@

A further means by which the efficiency of a propeller can be increased is through the reduction of the turbulence created by the blade-tips passing through the open (static) water at high speed.  Since the two sides of the blades have low and high pressure on them respectively, at the tips where the two faces meet, there there is a tendancy for water to `spill' from the high pressure side the the low pressure side, creating a vortex which leads to unwanted turbulence.  It is also around this vortex that blade tip cavitation tends to occur. By limiting the movement of water around the outside of the blade tip by having a tightly fitted shroud, losses due to this particulary type of tubulence, and the onset of this type of cavitation, or the vortices that occur even when cavitation does not, can be diminished.
